Collaborative Multioutput Gaussian Processes: \\ Supplementary Material

### 2014-03-19 paper/experiments.tex
sec 4 done

Since we are using stochastic optimization, the learn rates need to be chosen carefully.
We found that the rates used in \citep{hensmangaussian} also work well for our model.
Specifically, we used the learn rates of $0.01$ for the variational parameters, $1 \times 10^{-5}$ for the  covariance hyperparameters, and $1 \times 10^{-4}$ for the weights, noise hyperparameter, and inducing inputs.
We also included a momentum term of $0.9$ for all of the parameters except the variational parameters and the inducing inputs.
All of the experiments are executed on an Intel(R) Core(TM) i7-2600 3.40GHz CPU with 8GB of RAM using Matlab R2012a.

\subsection{TOY PROBLEM}
In this toy problem, two related outputs are simulated from the same latent function $sin(x)$ and corrupted by independent noises: $y_1(x) = sin(x) + \epsilon$ and $y_2(x) = -sin(x) + \epsilon$, $\epsilon \sim \Normal(0,0.01)$.
Each output is given 200 observations with missing values in the $(-7,-3)$ interval for the first output and the $(4,8)$ interval for the second output.
We used $Q = 1$ latent sparse process with squared exponential kernel, $h_1(x) = h_2(x) = 0$, and $M = 15$ inducing inputs for our model.

Figure \ref{fig:toy} shows the predictive distributions by our model (COGP) and independent GPs with stochastic variational inference (SVIGP, one for each output).
The locations of the inducing inputs are fixed and identical for both methods.
It is apparent from the figure that the independent GPs fail to predict the functions in the unobserved regions, especially for output 1.
In contrast, by using information from the observed intervals of one output to interpolate the missing signal of the other, COGP makes perfect prediction for both outputs.
This confirms the effectiveness of collaborative learning of sparse processes via the shared inducing variables. 
Also note that the inference procedure learned that the weights are $w_{11} = 1.07$ and $w_{21} = -1.06$ which accurately reflects the correlation between the two outputs.

@@

\subsection{FOREIGN EXCHANGE RATE PREDICTION}
The first real world application we consider is to predict the foreign exchange rate w.r.t the US dollar of the top 10 international currencies (CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN) and 3 precious metals (gold, silver, and platinum)\footnote{Data is available at http://fx.sauder.ubc.ca}. 
The setting of our experiment described here is identical to that in \citep{alvarez2010efficient}.
The dataset consists of all the data available for the 251 working days in the year of 2007.
There are 9, 8, and 42 days of missing values  for gold, silver, and platinum, respectively.
We remove from the data the exchange rate of CAD on days 50-100, JPY on day 100-150, and AUD on day 150-200.
Note that these 3 currencies are from very different geographical locations, making the problem more interesting. 
The 153 points are used for testing, and the remaining 3051 data points are used for training.
Since the missing data corresponds to long contiguous sections, the objective here is to evaluate the capacity of the model to impute the missing currency values based on other currencies.
%todo: batch size, learn rate (maybe at the beginning of experiments)

For preprocessing we normalized the outputs to have zero mean and unit variance.
Since the exchange rates are driven by a small number of latent market forces \citep[see e.g.][]{alvarez2010efficient}, we tried different values of $Q = {1,2,3}$ and selected $Q = 2$ which gave the best model evidence (ELBO).
We used the squared-exponential covariance function for the shared processes and the noise covariance function for the individual process of each output.
The inducing inputs $(M = 100, \text{per sparse process})$  are randomly selected from the training data and fixed throughout training.

@@

The real data and predictive distributions by our model are shown in Figure \ref{fig:fx}.
They exhibit similar behaviors to those by the convolved model with inducing kernels in \citep{alvarez2010efficient}.
In particular, both models perform better at capturing the strong depreciation of the AUD than the fluctuations of the CAD and JPY currency.
Our investigation of the dataset found that 4 other currencies (GBP, NZD, KRW, and MXN) also experienced the same trend during the days 150 - 200.
This information from these currencies was effectively used by the model to extrapolate the values of the AUD.

%todo: result for independent gps
We also report in Table \ref{tab:fx} the predictive accuracy of our model compared to the convolved GPs model with exact inference \citep[CGP,][]{alvarez-lawrence-nips-08} and independent GPs (IGP, one for each output).
Our model outperforms both of CGP and IGP in terms of the standardized mean squared error (SMSE).
CGP has lower negative log predictive density (NLPD), mainly due to the less conservative predictive variance of the exact CGP for the CAD currency.
% no std because there are 3 outputs but variance is small
For reference, the convolved GPs with approximation via the variational inducing kernels  \citep[CGPVAR,][]{alvarez2010efficient} has the SMSE of 0.2795 while the NLPD was not provided.
Training took only 10 minutes for our model compared to 1.4 hours for the full CGP model.

\setlength{\tabcolsep}{4pt}
\begin{table}[t]
\caption{Performance comparison on the foreign exchange rate dataset. Results are averaged of the 3 outputs over 5 repetitions. Smaller figures are better.}
\label{tab:fx}
\begin{center}
\begin{tabular}{ccc}
\toprule
\textbf{METHOD} & \textbf{SMSE} & \textbf{NLPD} \\ \hline
COGP  & \textbf{0.2125} & -0.8394 \\
CGP & 0.2427 & \textbf{-2.9474} \\
IGP & 0.5996 & 0.4082 \\
CGPVAR & 0.2795 & NA \\
%ICM & 0.3927 &
\bottomrule
\end{tabular}
\end{center}
\end{table}

@@

For pre-processing we normalized the outputs to have zero mean and unit variance.
We used $Q = 2$ sparse processes with the squared exponential covariance function and individual processes with the noise covariance function.
The inducing inputs $(M = 200)$ are randomly selected from the training set and fixed throughout training.

\begin{figure*}
\centering
\begin{tabular}{ccc}
\includegraphics[scale=0.3]{figures/cogp-weatherCambermet.eps} &
\includegraphics[scale=0.3]{figures/cgp-weatherCambermet.eps} &
\includegraphics[scale=0.3]{figures/weatherCambermet.eps}
\\
\includegraphics[scale=0.3]{figures/cogp-weatherChimet.eps} &
\includegraphics[scale=0.3]{figures/cgp-weatherChimet.eps} &
\includegraphics[scale=0.3]{figures/weatherChimet.eps}
\end{tabular}
\caption{Real data and predictive distributions by our method (COGP, left figures), the convolved GP method with exact inference (CGP, middle figures), and full independent GPs (right figures) for the air temperature problem. The coding color scheme is the same as in Figure \ref{fig:toy}.}
\label{fig:weather}
\end{figure*}

The real data and the predictive distributions by our model, CGP with exact inference \citep{alvarez-lawrence-nips-08}, and independent GPs are shown in Figure \ref{fig:weather}.
It is clear that the independent GPs model is clueless in the test regions and thus simply uses the average temperature as its prediction.
For Cambermet, both COGP and CGP can capture the rising in temperature from the morning till the afternoon and the fall afterward.
%For Chimet, both models perform poorly but CGP more so as it falsely predicts  wild fluctuations that are non-existent in the data. 
The comparative performance of the models are summarized in Table \ref{tab:air}, from which it can be seen that our model outperforms CGP in terms of both SMSE and NLPD.
%All results are averaged of the 2 outputs over 5 repetitions.
It took 5 minutes on average to train our model compared to 3 hours of CGP with exact inference.

\setlength{\tabcolsep}{4pt}
\begin{table}[t]
\caption{Performance comparison on the air temperature dataset. Results are averaged of 2 outputs over 5 repetitions. }
\label{tab:air}
\begin{center}
\begin{tabular}{ccc}
\toprule
\textbf{METHOD} & \textbf{SMSE} & \textbf{NLPD} \\
\hline
COGP & \textbf{0.1077} & \textbf{2.1712} \\
CGP & 0.1125 & 2.2219 \\
IGP & 0.8944 & 12.5319 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

It is also worth noting the characteristics of the sparse processes learned by our model as they indeed correspond to different patterns in the data.
In particular, one process has the inverse lengthscale of 136 which captures the global increase in temperature during the training period while the other has the inverse lengthscale of 0.5 to model the local variations within a single day.

\subsection{ROBOT INVERSE DYNAMICS}
Our last experiment is with a dataset relating to an inverse dynamics model of a 7-degree-of-freedom anthropomorphic robot arm \citep{vijayakumar2000locally}.
The data consists of 48,933 mapping from a 21-dimensional input space (7 joints positions, 7 joint velocities, 7 joint accelerations) to the corresponding 7 joint torques.
%The problem is strongly nonlinear due to complex superpositions of sine and cosine functions in robot dynamics \citep{vijayakumar2000locally}.
%Furthermore, exploratory analysis using standard GP with automatic relevance determination suggests that all of the 21 dimensions are relevant.
It has been used in previous work \citep[see e.g.][]{rasmussen-williams-book,vijayakumar2000locally} but only for single task learning. 
In \citep{chai2008multi}, multitask learning of robot inverse dynamics was considered but on a different and much smaller dataset.

@@

Also from Table \ref{tab:robotarm}, it seems that optimizing the inducing inputs can lead to better performance than fixing them.
More importantly, the overhead in computation is slight, as demonstrated by the training time using the two options in Figure \ref{fig:time}.
For instance, the total training time is only 1.9 hours when learning with 500 inducing inputs compared to 1.6 hours when fixing them.
As this dataset is 21-dimensional, the small difference in training time confirms our analysis that the cost of optimizing the inducing inputs only scales sublinearly with the dimensionality of the problem.
Thus, for datasets of high dimension, learning of the inducing inputs is a practical option.

>>>>

Since we are using stochastic optimization, the learning rates need to be chosen carefully.
We found that the rates used in \cite{hensmangaussian} also work well for our model.
Specifically, we used the learning rates of $0.01$ for the variational parameters, $1 \times 10^{-5}$ for the  covariance hyperparameters, and $1 \times 10^{-4}$ for the weights, noise hyperparameter, and inducing inputs.
We also included a momentum term of $0.9$ for all of the parameters except the variational parameters and the inducing inputs.
All of the experiments are executed on an Intel(R) Core(TM) i7-2600 3.40GHz CPU with 8GB of RAM using Matlab R2012a.
%
\subsection{TOY PROBLEM}
In this toy problem, two related outputs are simulated from the same latent function $sin(x)$ and corrupted by independent noise: $y_1(x) = sin(x) + \epsilon$ and $y_2(x) = -sin(x) + \epsilon$, $\epsilon \sim \Normal(0,0.01)$.
Each output is given 200 observations with missing values in the $(-7,-3)$ interval for the first output and the $(4,8)$ interval for the second output.
We used $Q = 1$ latent sparse process with squared exponential kernel, $h_1(x) = h_2(x) = 0$, and $M = 15$ inducing inputs for our model.

Figure \ref{fig:toy} shows the predictive distributions by our model (COGP) and independent GPs with stochastic variational inference (SVIGP, one for each output).
The locations of the inducing inputs are fixed and identical for both methods.
It is apparent from the figure that the independent GPs fail to predict the functions in the unobserved regions, especially for output 1.
In contrast, by using information from the observed intervals of one output to interpolate the missing signal of the other, COGP makes perfect prediction for both outputs.
This confirms the effectiveness of collaborative learning of sparse processes via the shared inducing variables. 
Also we note that the inference procedure learned that the weights are $w_{11} = 1.07$ and $w_{21} = -1.06$ which accurately reflects the correlation between the two outputs.

@@

\subsection{FOREIGN EXCHANGE RATE PREDICTION}
The first real world application we consider is to predict the foreign exchange rate w.r.t the US dollar of the top 10 international currencies (CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN) and 3 precious metals (gold, silver, and platinum)\footnote{Data is available at http://fx.sauder.ubc.ca}. 
The setting of our experiment described here is identical to that in \citep{alvarez2010efficient}.
The dataset consists of all the data available for the 251 working days in the year of 2007.
There are 9, 8, and 42 days of missing values  for gold, silver, and platinum, respectively.
We remove from the data the exchange rate of CAD on days 50--100, JPY on day 100--150, and AUD on day 150--200.
Note that these 3 currencies are from very different geographical locations, making the problem more interesting. 
The 153 points are used for testing, and the remaining 3051 data points are used for training.
Since the missing data corresponds to long contiguous sections, the objective here is to evaluate the capacity of the model to impute the missing currency values based on other currencies.
%todo: batch size, learn rate (maybe at the beginning of experiments)

For preprocessing we normalized the outputs to have zero mean and unit variance.
Since the exchange rates are driven by a small number of latent market forces \citep[see e.g.][]{alvarez2010efficient}, we tried different values of $Q = {1,2,3}$ and selected $Q = 2$ which gave the best model evidence (ELBO).
We used the squared-exponential covariance function for the shared processes and the noise covariance function for the individual process of each output.
$M = 100$ inducing inputs (\text{per sparse process}) were randomly selected from the training data and fixed throughout training.

@@

The real data and predictive distributions by our model are shown in Figure \ref{fig:fx}.
They exhibit similar behaviors to those by the convolved model with inducing kernels in \citet{alvarez2010efficient}.
In particular, both models perform better at capturing the strong depreciation of the AUD than the fluctuations of the CAD and JPY currency.
Further analysis of the dataset found that 4 other currencies (GBP, NZD, KRW, and MXN) also experienced the same trend during the days 
150 -- 200.
This information from these currencies was effectively used by the model to extrapolate the values of the AUD.

%todo: result for independent gps
We also report in Table \ref{tab:fx} the predictive performance of our model compared to the convolved GPs model with exact inference \citep[CGP,][]{alvarez-lawrence-nips-08} and independent GPs (IGP, one for each output).
Our model outperforms both of CGP and IGP in terms of the standardized mean squared error (SMSE).
CGP has lower negative log predictive density (NLPD), mainly due to the less conservative predictive variance of the exact CGP for the CAD currency.
% no std because there are 3 outputs but variance is small
For reference, the convolved GPs with approximation via the variational inducing kernels  \citep[CGPVAR,][]{alvarez2010efficient} has the SMSE of 0.2795 while the NLPD was not provided.
Training took only 10 minutes for our model compared to 1.4 hours for the full CGP model.

\setlength{\tabcolsep}{4pt}
\begin{table}[t]
\caption{Performance comparison on the foreign exchange rate dataset. Results are averages of the 3 outputs over 5 repetitions. Smaller figures are better.}
\label{tab:fx}
\begin{center}
\begin{tabular}{ccc}
\toprule
\textbf{METHOD} & \textbf{SMSE} & \textbf{NLPD} \\ \hline
COGP  & \textbf{0.2125} & -0.8394 \\
CGP & 0.2427 & \textbf{-2.9474} \\
IGP & 0.5996 & 0.4082 \\
%CGPVAR & 0.2795 & NA \\
%ICM & 0.3927 &
\bottomrule
\end{tabular}
\end{center}
\end{table}

@@

For pre-processing we normalized the outputs to have zero mean and unit variance.
We used $Q = 2$ sparse processes with the squared exponential covariance function and individual processes with the noise covariance function.
$M=200$  inducing inputs  were randomly selected from the training set and fixed throughout training.
%
\begin{figure*}
\centering
\begin{tabular}{ccc}
\includegraphics[scale=0.3]{figures/cogp-weatherCambermet.eps} &
\includegraphics[scale=0.3]{figures/cgp-weatherCambermet.eps} &
\includegraphics[scale=0.3]{figures/weatherCambermet.eps}
\\
\includegraphics[scale=0.3]{figures/cogp-weatherChimet.eps} &
\includegraphics[scale=0.3]{figures/cgp-weatherChimet.eps} &
\includegraphics[scale=0.3]{figures/weatherChimet.eps}
\end{tabular}
\caption{Real data and predictive distributions by our method (COGP, left figures), the convolved GP method with exact inference (CGP, middle figures), and full independent GPs (right figures) for the air temperature problem. The coding color scheme is the same as in Figure \ref{fig:toy}.}
\label{fig:weather}
\end{figure*}

The real data and the predictive distributions by our model, CGP with exact inference \citep{alvarez-lawrence-nips-08}, and independent GPs are shown in Figure \ref{fig:weather}.
It is clear that the independent GPs model is clueless in the test regions and thus simply uses the average temperature as its prediction.
For Cambermet, both COGP and CGP can capture the rising in temperature from the morning till the afternoon and the fall afterward.
%For Chimet, both models perform poorly but CGP more so as it falsely predicts  wild fluctuations that are non-existent in the data. 
The comparative performance of the models are summarized in Table \ref{tab:air}, from which it can be seen that our model outperforms CGP in terms of both SMSE and NLPD.
%All results are averaged of the 2 outputs over 5 repetitions.
It took 5 minutes on average to train our model compared to 3 hours of CGP with exact inference.
%
\setlength{\tabcolsep}{4pt}
\begin{table}[t]
\caption{Performance comparison on the air temperature dataset. Results are averaged of 2 outputs over 5 repetitions. }
\label{tab:air}
\begin{center}
\begin{tabular}{ccc}
\toprule
\textbf{METHOD} & \textbf{SMSE} & \textbf{NLPD} \\
\hline
COGP & \textbf{0.1077} & \textbf{2.1712} \\
CGP & 0.1125 & 2.2219 \\
IGP & 0.8944 & 12.5319 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}
%
It is also worth noting the characteristics of the sparse processes learned by our model as they indeed correspond to different patterns in the data.
In particular, one process has the inverse lengthscale of 136 which captures the global increase in temperature during the training period while the other has the inverse lengthscale of 0.5 to model the local variations within a single day.

\subsection{ROBOT INVERSE DYNAMICS}
Our last experiment is with a dataset relating to an inverse dynamics model of a 7-degree-of-freedom anthropomorphic robot arm \citep{vijayakumar2000locally}.
The data consists of 48,933 mapping from a 21-dimensional input space (7 joints positions, 7 joint velocities, 7 joint accelerations) to the corresponding 7 joint torques.
%The problem is strongly nonlinear due to complex superpositions of sine and cosine functions in robot dynamics \citep{vijayakumar2000locally}.
%Furthermore, exploratory analysis using standard GP with automatic relevance determination suggests that all of the 21 dimensions are relevant.
It has been used in previous work \citep[see e.g.][]{rasmussen-williams-book,vijayakumar2000locally} but only for single task learning. 
\citet{chai2008multi} considered multitask learning of robot inverse dynamics  but on a different and much smaller dataset.

@@

Also from Table \ref{tab:robotarm}, it seems that optimizing the inducing inputs can lead to better performance than fixing them.
More importantly, the overhead in computation is small, as demonstrated by the training time using the two options in Figure \ref{fig:time}.
For instance, the total training time is only 1.9 hours when learning with 500 inducing inputs compared to 1.6 hours when fixing them.
As this dataset is 21-dimensional, the small difference in training time confirms our analysis that the cost of optimizing the inducing inputs only scales sublinearly with the dimensionality of the problem.
Thus, for datasets of high dimension, learning of the inducing inputs is a practical option.
